Linear stability of magnetized massive protoplanetary disks

### 2014-04-05 paper/results_azi.tex
tex updates

\section{Effect of an azimuthal field}\label{result3}
In this section we use the setup of case 2 described in
\S\ref{linear}, and examine the effect of an azimuthal field so that
$B_y\neq 0$,   
parameterized by $\epsilon \equiv B_y/B_z$. However, we continute to
use $B_z$ for normalizations and $\beta$ is associated with the
vertical Alfven speed. We also extend the previous calculations to the
full disk $z\in[-Z_s,Z_s]$,  which allows us to compare the effect of
self-gravity on MRI modes with different symmetries across the
midplane.

\subsection{Ideal disks with MRI} 
We consider an isothermal disk with $Q=0.2$ ($Q_\mathrm{2D}=~0.72$) and
$\beta=10$ in the limit of ideal MHD ($\Lambda_0=100$). Gravitational
instability is not expected because Fig. \ref{compare_growth3} shows
that even for $Q=0.18$, GI is surpressed for $\beta \lesssim 15$. %We
%consider azimuthal field strengths with $\epsilon\in[0,3]$.

Fig. \ref{compare_growth3_tilted} show MRI growth rates for
$B_y/B_z=0,\,1,\,2$ and $3$. We divide the modes into two catagories
depending on the extremum of magnetic energy at the midplane. 
The top panel are modes where $E_m$ has a local minimum at $z=0$ and 
the bottom panel are modes where $E_m$ has a local maximum at $z=0$. 
The latter set of modes were excluded in the previous sections by
midplane boundary conditions. 
%The top panel are modes with even midplane symmetry: 
%$W^\prime(0)= 0$ for $\epsilon\to0$, and the bottom panel are modes with odd midplane 
%symmetry: $W(0) = 0$ for $\epsilon\to0$. 
We also plot growth rates computed in the Cowling
approximation. As expected,  $\avg{E_g}<\avg{E_m}$, so none of the
modes are energetically dominated by self-gravity.

%no GI for odd modes 
%general stabilization of MRI

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_even.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_odd.ps} 
  \caption{Growth rates in massive isothermal disks with $Q=~0.2$ and
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation, while
    the solid curves are computed from the Cowling approximation. The
    top panel correspond to even modes and the
    bottom panel correspond to odd modes, which have
    $W^\prime(0)=0$ and $W(0)=0$, respectively, for $\epsilon\to0$.
    \label{compare_growth3_tilted}}
\end{figure}

%We consider $\epsilon\equiv B_y/B_z \in[0,3]$. The

Consider first modes in the top panel of 
Fig. \ref{compare_growth3_tilted}. As with previous results,  
self-gravity destabilizes modes with  $k_xH\gtrsim O(1)$. Consequently, the
cut-off wavenumber is larger when SG is included. %get smaller scale
                                %MRI 
Destablization is most effective for purely vertical fields: with
$\epsilon=0,\, k_xH\simeq 1.4$, SG increases the growth rate by $\sim
30\%$. For $B_y=0$ we find the density perturbation $W(z)$ is an even
function. Although these modes are fundamentally magnetic, this is consistent with
\cite{goldreich65a}, who showed that SG can only destabilize 
symmetric density perturbations.  
With increasing $\epsilon$, we find $W$ deviates from an even
function. %,% $|W(0)|$ decreases and $\max{(|W|)}$ 
%moves off of the mid-plane.    
%density pert at the midplane goes down
%We expect this to reduce the destabilization
%effect of SG,  
Together with the increased total magnetic pressure with 
$\epsilon$, destabilization by SG weakens. 
Thus, the Cowling approximation becomes increasingly good with stronger
$B_y$ for these modes.

%This destabilization effect weakens with
%$B_y$ because the total magnetic pressure increases, which generally
%opposes GI, although the modes are fundamentally associated with MRI.

The modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. For $B_y=0$ we find $W$ is odd, and
self-gravity has no effect. When $B_y>0$, $W$ deviates from an
odd function and the mid-plane density perturbation $|W(0)|$
increases. SG is stabilizing for these modes at all wavelengths, and
is most effective at $k_xH =  O(1)$.  
%destabilizing effect is absent  
%for $\epsilon=0,\, k_xH\simeq 1.2$, SG decreases the growth rate by $\sim
%13\%$.

Fig. \ref{result_tilted} show eigenfunctions for $\epsilon=3$ and
$k_xH=1.1$ with and without the Cowling approximation. 
%These differ  
%from the even modes considered above in that the magnetic energy
%maximize at $z=0$ (cf. Fig. \ref{mri_massive_resis}).  
SG significantly enhances the midplane density perturbation, making the
gravitational potential energy comparable to the magnetic energy, 
which becomes more confined near the mid-plane.

@@

%strong field, compressibility 
To interpret the above result for modes with magnetic energy
concentrated at the midplane, we note that compressibility affects the
MRI in the presence of an azimuthal field even in a
non-self-gravitating disk. If the perturbed disk remains in vertical
hydrostatic equilibrium, then  
\begin{align} 
  |W|\sim \frac{B_y}{\mu_0\rho}|\dby|,
\end{align}
to order of magnitude in a non-SG disk. Thus a strong azimuthal field can 
cause a large density perturbation \citep{pessah05}. We checked that
for the modes in Fig. \ref{result_tilted}, vertical velocities are 
small, $|\dvz|/\left(|\dvx|^2+|\dvy|^2\right)^{1/2} \lesssim 0.2$.

Compressibilty is enhanced by an azimuthal field, which has a
stabilizing effect on the MRI \citep{kim00}. %because...?
This effect is significant for $\epsilon=3$ because the azimuthal
Alfven speed is sonic.    
%maybe blaes 
Fig. \ref{result_tilted} indicates that self-gravity further enhances
compressibility, and therefore stabilization. We suspect this is
overwhelmed by the destabilization effect of SG because the density
perturbation has an anit-symmetric component.

%compressibility -> pressure restoring force

%The results for even and odd modes are do not contradict each
%other. Self-gravity tends to be destabilizing, but this requires the
%density perturbation to be symmetric, as is the case for the even
%modes. 
%For the MRI however, compressibility has a stabilizing effect

%scale separation for odd modes? because modes with kx=1 will be
%quenched

% for $\beta=10$ and
%$\epsilon=3$. 
%because magnetic energy is concentrated near midplane    
%magnetic pressure, need thermal pressure 
%This
%can subsequently be amplitfied by its self-gravity. 
%That is, compressibility is enhanced by both
%For $\epsilon=3$ and $\beta=10$ the azimuthal Alven speed is
%sonic. Compressibility becomes important for strong toroidal
%fields. If the system remains in approximately vertical hydrod

%\subsection{Ideal disks with GI (TBC)}

\subsection{Resistive disks with GI}
Here we consider an isothermal, resistive disk with $Q=0.18$,
$\Lambda_0=0.1$ and $\beta=100$. This system permits GI and MRI. 
%GI is permitted, and has a similar
%growth rate as the  
%MRI. 
Fig. \ref{compare_growth3_tilted_resis} show growth rates for
$\epsilon=0,\,1$ and $2$. For $B_y=0$, MRI and GI are decoupled except 
for a narrow range of $k_x$ in which the lower MRI modes transitions to
GI. Notice that the upper MRI modes intersect the GI branch. There is no
interaction because the upper MRI modes have anti-symmetric $W(z)$ whereas
the GI modes have symmetric $W(z)$.

%GI, and there are MRI modes available on that scale

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_resis_ep0.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep1.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep2.ps}
  \caption{Growth rates in isothermal resistive disks with $Q=0.18$
    ($Q_\mathrm{2D}=0.67$), $\beta=100$ and $\Lambda_0=0.1$. For
    $B_y=0$, the upper MRI modes have anti-symmetric density
    perturbations, $W(0) = 0$; while the lower MRI modes have symmetric
    density perturbations, $W^\prime(0) = 0$. For $B_y/B_z=2$ the
    overstable modes have non-zero real frequencies. 
    \label{compare_growth3_tilted_resis}}
\end{figure}

Introducing $B_y = B_z$ leads to an exchange in the mode
characters. For $k_xH\lesssim 0.9$ the modes on the two MRI branches
are similar to the vertical field case. However, for $k_xH\gtrsim0.9$ 
the upper MRI mode transitions to GI, for which $E_m(0)$ is a
minimum; and the lower MRI mode has $E_m(0)$ being a maximum. 
%We
%believe this interchange is enabled by the presence of GI. 
%interchange possible because of GI branch

%In both cases where the gravitational energy is significant, the
%density perturbation is symmetric.

Increasing the azimuthal field further to $B_y=2B_z$ we find
overstable MRI modes \citep{gammie96} with non-negligible real
frequencies. An example is shown in
Fig. \ref{result_tilted_overstable}. Notice the density/potential
perturbation is off-set from the midplane. This is not possible for
pure GI \citep{goldreich65a}. Thus, these overstable MRI modes indeed
become self-gravitating, before being stabilized. 
%these may lead to significant vertical motions

Notice in Fig. \ref{compare_growth3_tilted_resis} the disappearence of
magnetic modes between $0.8\lesssim k_xH\lesssim 1.5$ as $B_y$ is
increased. For $B_y=2B_z$, MRI and GI operate at distinct radial
scales. This implies that perturbations unstable to GI cannot develop
MRI.

%This may lead to 
%the lower MRI mode transitions to the top MRI mode

@@

%\subsection{GI}

>>>>

\section{Effect of an azimuthal field}\label{result3}
In this section we use the setup of case 2 described in
\S\ref{linear}, and examine the effect of an azimuthal field so that
$B_y\neq 0$,   
parameterized by $\epsilon \equiv B_y/B_z$. However, we continute to
use $B_z$ for normalizations and $\beta$ is associated with the
vertical Alfven speed. We also extend the previous calculations to the
full disk $z\in[-Z_s,Z_s]$,  which allows us to compare the effect of
self-gravity on MRI modes with different symmetries across the
midplane. We use an isothermal disk throughout.

\subsection{Ideal disks with MRI} 
We consider disk with $Q=0.2$ ($Q_\mathrm{2D}=~0.72$) and
$\beta=10$ in the limit of ideal MHD ($\Lambda_0=100$). Gravitational
instability is not expected because Fig. \ref{compare_growth3} shows
that even for $Q=0.18$, GI is surpressed for $\beta \lesssim 15$.

Fig. \ref{compare_growth3_tilted} show MRI growth rates for
$B_y/B_z=0,\,1,\,2$ and $3$. We divide the modes into two catagories
depending on the extremum of magnetic energy at the midplane. 
The top panel are modes where $E_m$ has a local minimum at $z=0$ and 
the bottom panel are modes where $E_m$ has a local maximum at $z=0$. 
The latter set of modes were excluded in the previous sections by
midplane boundary conditions. 
We also plot growth rates computed in the Cowling
approximation. As expected,  $\avg{E_g}<\avg{E_m}$, so none of the
modes are energetically dominated by self-gravity.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_even.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_odd.ps} 
  \caption{Growth rates in massive isothermal disks with $Q=~0.2$ and
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation, while
    the solid curves are computed from the Cowling approximation. 
    For $B_y=0$, modes in top and bottom panels 
    $W^\prime(0)=0$ and $W(0)=0$, respectively.
    \label{compare_growth3_tilted}}
\end{figure}

Consider first modes in the top panel of 
Fig. \ref{compare_growth3_tilted}. As with previous results,  
self-gravity destabilizes modes with  $k_xH\gtrsim O(1)$. Consequently, the
cut-off wavenumber is larger when SG is included. 
Destablization is most effective for purely vertical fields: with
$\epsilon=0,\, k_xH\simeq 1.4$, SG increases the growth rate by $\sim
30\%$. For $B_y=0$ we find the density perturbation $W(z)$ is an even
function. Although these modes are fundamentally magnetic, this is consistent with
\cite{goldreich65a}, who showed that SG can only destabilize 
symmetric density perturbations.  
With increasing $B_y$, we find $W$ deviates from an even
function. 
Together with the increased total magnetic pressure with 
$B_y$ (since $B_z$ is fixed), destabilization by SG weakens. 
Thus, the Cowling approximation becomes increasingly good with stronger
$B_y$ for these modes.

The modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. For $B_y=0$ we find $W$ is odd, and
self-gravity has no effect. When $B_y>0$, $W$ deviates from an
odd function and the midplane density perturbation $|W(0)|$
increases. SG is stabilizing for these modes at all wavelengths, and
is most effective at $k_xH =  O(1)$.

Fig. \ref{result_tilted} show eigenfunctions for $\epsilon=3$ and
$k_xH=1.1$ with and without the Cowling approximation.
SG significantly enhances the midplane density perturbation, making the
gravitational potential energy comparable to the magnetic energy, 
which becomes more confined near the midplane.

@@

To interpret the above result for modes with magnetic energy
concentrated at the midplane, we note that compressibility affects the
MRI in the presence of an azimuthal field even in a
non-self-gravitating disk. If the perturbed disk remains in vertical
hydrostatic equilibrium, then  
\begin{align} 
  |W|\sim \frac{B_y}{\mu_0\rho}|\dby|,
\end{align}
to order of magnitude in a non-SG disk. Thus a strong azimuthal field can 
cause a large density perturbation \citep{pessah05}. We checked that
for the modes in Fig. \ref{result_tilted}, vertical velocities are 
small, $|\dvz|/\left(|\dvx|^2+|\dvy|^2\right)^{1/2} \lesssim 0.2$.

Compressibilty is enhanced by an azimuthal field, which is 
stabilizing for the MRI \citep{kim00}. %because...?
This effect is significant for $\epsilon=3$ because the azimuthal
Alfven speed is sonic.      
%maybe blaes 
Fig. \ref{result_tilted} indicates that self-gravity further enhances
compressibility, and therefore stabilization. We suspect this is
overwhelmed by the destabilization effect of SG, because the density
perturbation has an anti-symmetric component.

\subsection{Resistive disks with GI}
Here we examine a resistive disk which permits MRI and GI by setting $Q=0.18$,
$\Lambda_0=0.1$ and $\beta=100$. 
Fig. \ref{compare_growth3_tilted_resis} show growth rates for
$\epsilon=0,\,1$ and $2$. For $B_y=0$, MRI and GI are decoupled except 
for a narrow range of $k_x$ in which the lower MRI modes transitions to
GI. Notice that the upper MRI modes intersect the GI branch. There is no
interaction because the upper MRI modes have anti-symmetric $W(z)$ whereas
the GI modes have symmetric $W(z)$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_resis_ep0.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep1.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep2.ps}
  \caption{Growth rates in isothermal resistive disks with $Q=0.18$
    ($Q_\mathrm{2D}=0.67$), $\beta=100$ and $\Lambda_0=0.1$. For
    $B_y=0$, the upper and lower MRI modes have anti-symmetric and
    symmetric density perturbations, corresponding to $W(0)=0$ and
    $W^\prime(0)=0$, respectively.  
    %anti-symmetric density
    %    perturbation $W(0) = 0$ and symmetric density perturbations,
    %    $W^\prime(0) = 0$.  
    For $B_y/B_z=2$ the
    overstable modes have non-zero real frequencies. 
    \label{compare_growth3_tilted_resis}}
\end{figure}

Introducing $B_y = B_z$ leads to an exchange in the mode
characters. For $k_xH\lesssim 0.9$ the modes on the two MRI branches
are similar to the vertical field case. However, for $k_xH\gtrsim0.9$ 
the upper MRI mode transitions to GI, for which $E_m(0)$ is a
minimum; and the lower MRI mode has $E_m(0)$ being a maximum. We find
all perturbations with  $k_xH\gtrsim0.9$ have symmetric $W(z)$. 
%small scale perts are all even

Increasing the azimuthal field further to $B_y=2B_z$ we find
overstable MRI modes  with non-negligible real
frequencies \citep{gammie96}. An example is shown in
Fig. \ref{result_tilted_overstable}. Notice the density/potential
perturbation is off-set from the midplane. This is not possible for
pure GI \citep{goldreich65a}. Thus, these overstable MRI modes indeed
become self-gravitating, before being stabilized. 
%these may lead to significant vertical motions

Notice also in Fig. \ref{compare_growth3_tilted_resis} the disappearence of
magnetic modes between $0.8\lesssim k_xH\lesssim 1.5$ as $B_y$ is
increased. For $B_y=2B_z$, MRI and GI are again independent, but
because they operate at distinct radial scales. This implies that
perturbations unstable to GI cannot develop MRI.

@@

### 2014-04-05 paper/results_gi.tex
tex updates

\subsection{Co-existence of MRI and GI} 
Fig. \ref{compare_growth3} show growth rates for modes with $k_xH=1$
as a function of $\beta$ in disks with $Q=0.18$, $Q=0.14$ and
$Q=0.12$.   
All three cases display distinct GI modes (red/brown branch). The GI 
growth rates are $\gamma\simeq 0.25\Omega,\,0.6\Omega,\,0.8\Omega$ for
$Q=0.18,\,0.14,\,0.12$, respecitvely. GI is stabilized by magnetic
pressure for sufficiently small $\beta$. The critical
field strength increases with increasing self-gravity \citep{nakamura83}. 
For $Q=0.18$, GI is stabilized for $\beta \lesssim 15$. Nevertheless, 
the MRI branch for $\beta < 15$ becomes self-gravitating, so that
density perturbations will still develop, even though GI does not
formally operate.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0cm]{figures/compare_growth3_kx1_Q0d18.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d14.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d12.ps} 
  \caption{Growth rates for modes with $k_xH=1$ in an isothermal disk with
    $Q=0.18$ (top), $Q=0.14$ (middle) and $Q=0.14$ (bottom). The
    colorbar measures the importance of  
    self-gravity with $\tau$.   
    \label{compare_growth3}}
\end{figure}

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=0.18$
where the GI branch approaches an MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. In fact, following the red curve to smaller
$\beta$ indicates GI transitions to MRI. The
`gaps' in the GI and MRI branches for $Q=0.18$ and $Q=0.12$ may be due
to the phenomenon of avoided crossing, as seen in stars
\citep[e.g.][]{aizenman77} and accretion tori
\citep[e.g.][]{christo93}, where physically distinct modes approach
one another in frequency and exchange character. However, we cannot
exclude the possibility  that some modes may have been missed in a 
numerical search of eigenfrequencies.

Thus, our results do not rigorously prove that the GI and MRI branches
do not intersect.  Nevertheless, the continous variation of $\tau$
strongly suggest that unstable modes can transition smoothly from 
MRI to GI and vice versa, especially at low $\beta$.

@@

We also plot in Fig. \ref{compare_growth3_Q01d2} growth rates obtained 
from the Cowling approximation, which isolates MRI; and that from a
high-resistivity run, which isolates GI by allowing the field lines to
slip through the fluid. We refer to these as  
pure MRI and pure GI, respectively.  
For $k_xH\lesssim 0.2$, growth rates are equal to
those on the pure MRI and pure GI branches. That is, MRI and GI
operate independently until their growth rates become equal as a
function of $k_x$.

The dispersion relation $\gamma(k_x)$ deviates from the
pure GI/MRI curves with increasing $k_x$, implying stronger interaction between
magnetic and density perturbations. Comparing pure GI (dashed line)
and the gravitationally-dominated portions of $\gamma(k_x)$ shows that 
inclusion of magnetic field stabilizes high-$k_x$ pure GI. (Note also
the slight decrease in the most unstable $k_x$.) %Since axisymmetric GI
%is unaffected by shear, 
This stabilization is due to magnetic pressure \citep{lizano10}, which
is consistent with stabilizing the small wavelength GI only.

%3D geometry doesn't affect 
%axisymmetric GI doesn't care about shear 
%
%two legs of GI are not symmetric 
%MRI become self-gravitating
%GI with increasing magnetization

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity increases MRI growth
rates at large $k_x$, but the effect is small. %can be used for code
                                %test 
Note that this 
destabilization is through the linear response, rather than through
the background stratification (which is stabilizing). 
%For the $k_x$ in the GI
%interval, MRI is weakly stabilized by self-gravity.

%magnetic pressure, small scale more effective 
%, reflected in the full range of $\tau$ displayed by the
%modes. 
%mode classification becomes difficult. 
%the point is: dispersion relation cannot be estimated by computing
%the MRI and GI separately

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth3_Q0d12_B20.ps}  
  \caption{Growth rates of unstable modes in the massive isothermal
    disk with $Q=0.12$ and $\beta=20$, as a function of the horizontal
    wavenumber $k_x$. The colorbar measures the importance of
    self-gravity. The solid line corresponds to MRI modes in the
    Cowling approximation. The dashed line corresponds to pure GI
    modes, obtained by including a high resisitivity in the full
    problem. 
%    {\bf REDO FIGURE}
    \label{compare_growth3_Q01d2}}
\end{figure}

%implications for simulations -> need to resolve small-scales

%Fig. \ref{compare_growth3_Q01d2} again shows smooth
%transitions between MRI and GI. This means the MRI and GI do indeed
%interact, but

%In particular, modes with $1.5 \lessim
%k_xH\lesssim2$ is neither pure

%demonstrates that mode classification
%becomes ambigious when both GI and MRI are present.

%\subsection{Layered disks}
%Can a disk with a dense, resistive midplane and a highly conducting
%atmosphere  be subject to both MRI and GI but at different heights?

%GI in the midplne and MRI in the surface?

>>>>

\subsection{Co-existence of MRI and GI} 
Fig. \ref{compare_growth3} show growth rates for modes with $k_xH=1$
as a function of $\beta$ in disks with $Q=0.18$, $Q=0.14$ and
$Q=0.12$.   
All three cases display distinct GI modes (red/brown branch). The GI 
growth rates are $\gamma\simeq 0.25\Omega,\,0.6\Omega,\,0.8\Omega$ for
$Q=0.18,\,0.14,\,0.12$, respecitvely. GI is stabilized by magnetic
pressure for sufficiently small $\beta$. The critical
field strength for stabilizating GI increases with increasing
self-gravity, consistent with \cite{nakamura83}.   
For $Q=0.18$, GI is stabilized for $\beta \lesssim 15$. Nevertheless, 
the MRI branch for $\beta < 15$ becomes self-gravitating, so that
density perturbations still grow, even though GI does not
formally operate.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0cm]{figures/compare_growth3_kx1_Q0d18.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d14.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d12.ps} 
  \caption{Growth rates for modes with $k_xH=1$ in isothermal ideal
    disks with $Q=0.18$ (top), $Q=0.14$ (middle) and $Q=0.14$ (bottom). The
    colorbar measures the importance of  
    self-gravity by $\tau$.   
    \label{compare_growth3}}
\end{figure}

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=~0.18$
where the GI branch approaches a MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. In fact, following the red curve to smaller
$\beta$ indicates GI transitions to MRI. The
`gaps' in the GI and MRI branches for $Q=0.18$ and $Q=0.12$ may be due
to the phenomenon of avoided crossing, as seen in stars
\citep[e.g.][]{aizenman77} and accretion tori
\citep[e.g.][]{christo93}, where physically distinct modes approach
one another in frequency and exchange character. However, we cannot
exclude the possibility  that some modes may have been missed in a 
numerical search of eigenfrequencies.  
%maybe ogilvie?

Thus, our results do not rigorously prove that the GI and MRI branches
do not intersect.  Nevertheless, the continous variation of $\tau$
strongly suggest that unstable modes can transition smoothly from 
MRI to GI and vice versa, especially at low $\beta$.     
%no unique transition

@@

We also plot in Fig. \ref{compare_growth3_Q01d2} growth rates obtained 
from the Cowling approximation, which isolates MRI; and that from a
high-resistivity run, which isolates GI by allowing the field lines to
slip through the fluid. We refer to these as  
pure MRI and pure GI, respectively.  
For $k_xH\lesssim 0.7$, growth rates are equal to
those on the pure MRI and pure GI branches. That is, MRI and GI
operate independently until their growth rates become equal as a
function of $k_x$.

The dispersion relation $\gamma(k_x)$ deviates from the
pure GI/MRI curves with increasing $k_x$, implying stronger interaction between
magnetic and density perturbations. Comparing pure GI (dashed line)
and the gravitationally-dominated portions of $\gamma(k_x)$ shows that 
inclusion of magnetic field stabilizes high-$k_x$ pure GI. (Note also
the slight decrease in the most unstable $k_x$.) 
This stabilization is due to magnetic pressure \citep{lizano10}, which
is consistent with stabilizing the small wavelength GI only.

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity increases MRI growth
rates at large $k_x$. This effect is small but noticeable, which can
be used as a code test for non-linear simulations. Note that this
destabilization is through the linear response, rather 
than through the background stratification (which is stabilizing).

%smooth transitions, mode exchange

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth3_Q0d12_B20.ps}  
  \caption{Growth rates of unstable modes in the massive isothermal
    disk with $Q=0.12$ and $\beta=20$, as a function of the horizontal
    wavenumber $k_x$. The colorbar measures the importance of
    self-gravity by $\tau$. The solid line corresponds to MRI modes in the
    Cowling approximation. The dashed line corresponds to pure GI
    modes, obtained by including a high resisitivity in the full
    problem. 
    \label{compare_growth3_Q01d2}}
\end{figure}
